Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-18 latex/main.tex
Add section on time travellers

\subsubsection*{Handling outlier samples}

Although most samples can be explained by a small number of mutations in the ARG,
there are two types of outlier samples which have unusually high numbers of mutations:
``time travelers'' and ``saltation'' lineages.
Time traveling samples are caused by errors in their collection dates.
For example, if a sample is actually collected 10 months after its reported collection date,
then 20 mutations (instead of, say, 0 or 1) are needed to explain it
using the ARG available on its collection date,
assuming a substitution rate of 2 events per genome per month.
Samples from ``saltation'' lineages (particularly the major VOCs) are characterized
by high numbers of mutations by the time of their detection.
Both types of outlier samples are poorly fit to the ARG when they appear on their collection dates.
To handle these outlier samples, we introduce three mechanisms:
(1) HMM cost-based rejection,
(2) retrospective sample grouping, and
(3) unconditional seeding.

@@

Of the 929 recombination events in the ARG,
this measure identifies 543 as artifactual,
leaving 354 QC-passing events (Figure~\ref{fig:recombinant_qc}).
Evidence that this filter effectively removed artifacts can be seen
in the clear association between the AmpliSeq V1 kit and
recombination events that we identified as artifactual.

Furthermore,
the illustrative copying patterns in Figure~\ref{fig:copying_pattern_examples}
depict different types of artefacts that were successfully removed.
Q1 shows an example of a recombinant caused by an MNS (A507T, T508C, G509A),
which is a documented sequencing error due to incomplete read trimming \cite{DeMaio2024}.
Q4 shows an example of a recombinant caused by a Delta lineage-defining deletion.
The vast majority of the QC-failing recombinants has this deletion, and
many cases are affected by a known amplicon dropout caused by this particular deletion.
Q3 shows an example artefact caused by a combination of an MNS and a deletion.

>>>>

\subsubsection*{Filtering time travellers}
Time-travelling samples---those in which the recorded collection date differs
substantially from their true collection date---pose significant problems
for sc2ts, and are present at a significant frequency. For example, of the 
43 Scorpio designations present in the Viridian dataset studied, 32 have 
samples present with sampling dates in 2020, including lineages that do not
arise until late 2022. Inserting these time-travellers into the ARG 
at the claimed sampling date causes major issues, and they must therefore be 
filtered out. We use the simple approach of filtering out samples that 
exceed a given ``HMM cost'', i.e., requiring too many mutations or
recombination breakpoints to plausibly represent the steady accumulation
of diversity expected with dense sampling. 
The HMM cost for the Viterbi solution for a sample under the LS model
is $k$ times the number of recombination events plus the number of 
mismatches, where $k$ is the recombination penalty (here, $k=4$).
After some experimentation, we set the recombination penalty to 7, 
disallowing samples that require 8 or more mutations (or, e.g., 1 recombination
and 4 mutations, or 2 recombinations) from immediate inclusion (see next
section) in the ARG. This approach has the benefit of filtering
samples enriched for errors as well as time-travellers.

\subsubsection*{Handling outlier samples}
Although most samples can be explained by a small number of mutations in the ARG,
there are two types of outlier samples which result in unusually high numbers of mutations:
``time travelers'' and ``saltation'' lineages.
Time traveling samples are caused by errors in their collection dates.
For example, if a sample is actually collected 10 months after its reported collection date,
then 20 mutations (instead of, say, 0 or 1) are needed to explain it
using the ARG available on its collection date,
assuming a substitution rate of 2 events per genome per month.
Samples from ``saltation'' lineages (particularly the major VOCs) are characterized
by high numbers of mutations by the time of their detection.
Both types of outlier samples are poorly fit to the ARG when they appear on their collection dates.
To handle these outlier samples, we introduce three mechanisms:
(1) HMM cost-based rejection,
(2) retrospective sample grouping, and
(3) unconditional seeding.

@@

Of the 929 recombination events in the ARG,
this measure identifies 543 as artifactual,
leaving 354 QC-passing events (Figure~\ref{fig:recombinant_qc}).

Furthermore,
the example copying patterns in Figure~\ref{fig:copying_pattern_examples}
illustrate different types of artefacts removed using this filter.
Q1 shows an example of a recombinant caused by an MNS (A507T, T508C, G509A),
which is a documented sequencing error due to incomplete read trimming \cite{DeMaio2024}.
Q4 shows an example of a recombinant caused by a Delta lineage-defining deletion.
The vast majority (XX, or XX\%) of the QC-failing recombinants has this deletion, and
many cases are affected by a known amplicon dropout caused by this particular deletion.
Q3 shows an example artefact caused by a combination of an MNS and a deletion.

Evidence that this filter effectively removed artifacts can be seen
in the clear association between the AmpliSeq V1 kit and
recombination events that we identified as artifactual.
